\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2024-03-04 misc/JSS article/jss-article-rnw/article.Rnw
Minors + large footnote on benchmarking.

%% - \Abstract{} almost as usual
\Abstract{
\pkg{collapse} is a large C/C++-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and programming efficiency. It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g., \class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling its seamless integration with large parts of the \proglang{R} ecosystem. This article introduces the package's key components and design principles in a structured way, supported by a rich set of examples. A small benchmark demonstrates its computational performance.
%  This short article illustrates how to write a manuscript for the
%  \emph{Journal of Statistical Software} (JSS) using its {\LaTeX} style files.
%  Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
%  it is recommended to keep the {\LaTeX} code as simple as possible,
%  i.e., avoid inclusion of packages/commands that are not necessary.
%  For outlining the typical structure of a JSS article some brief text snippets
%  are employed that have been inspired by \cite{Zeileis+Kleiber+Jackman:2008},
%  discussing count data regression in \proglang{R}. Editorial comments and
%  instructions are marked by vertical bars.
}

@@

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}}\footnote{Website: https://sebkrantz.github.io/collapse/} is a large C/C++-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions. Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently. It also offers vectorizations for advanced operations such as weighted statistics (including mode and quantiles), functions and classes for fully indexed (time-aware) computations on time series and panel data, recursive (list-processing) tools to deal with nested data and advanced descriptive statistical tools. This functionality is supported by efficient algorithms for intensive operations like grouping, unique values, matching, ordering, etc., tailored to \proglang{R}'s data structures and powerful data manipulation functions. The package also supplies many features for memory efficient \proglang{R} programming, such as data transformation and math by reference and aversion of logical vectors. \pkg{collapse} is class-agnostic, i.e., it provides most statistical operations for atomic vectors, matrices, and data frames/lists, and seamlessly supports key variants of these objects used in the \proglang{R} ecosystem (e.g., \emph{tibble}, \emph{data.table}, \emph{sf}, \emph{xts}, \emph{pdata.frame}). It is globally and interactively configurable, which includes setting different defaults for key function arguments (such as \code{na.rm} arguments to statistical functions, default \code{TRUE}), and modifying the package namespace itself.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse}, but can be interactively modified to overwrite key functions like \code{unique}, \code{match}, \code{\%in\%}, \code{table}, \code{subset}, \code{mutate}, \code{summarise} etc. with much faster \pkg{collapse} equivalents.} \newline

What is the purpose of combining all of this in a package? The short answer is to make computations in \proglang{R} as flexible and powerful as possible. The more elaborate answer is to (1) facilitate complex data transformation, exploration, and computing tasks in \proglang{R}; (2) increase performance and parsimony by avoiding \proglang{R}-level repetition\footnote{Such as applying \proglang{R} functions across columns or split-apply-combine computing to apply functions across groups or other divisions of data.}; (3) increase the memory efficiency and flexibility of \proglang{R} programs\footnote{E.g., by avoiding object conversions and the need for certain classes to do certain things, such as converting to data frame or \class{data.table} to do something "by groups" and then convert back to matrix to continue with linear algebra, and in general to reduce the need for metaprogramming.}; and (4) to create a new foundation package for statistics and data manipulation in \proglang{R} that implements successful ideas developed in the \proglang{R} ecosystem and other programming environments such as \proglang{Python} or \proglang{STATA} \citep{STATA}, including some new ideas, in a stable, high performance, and broadly compatible manor.\footnote{Examples of such ideas are \pkg{tidyverse} syntax, vectorized aggregations (\pkg{data.table}), data transformation by reference (\proglang{Python}, \pkg{pandas}), vectorized and verbose joins (\pkg{polars}, \proglang{STATA}), indexed time series and panel data (\pkg{xts}, \pkg{plm}), summary statistics for panel data (\proglang{STATA}), reshaping labelled data (myself) etc...} \newline

@@

\pkg{collapse} relates to and integrates key elements from these projects. It offers \pkg{tidyverse}-like data manipulation at the speed and stability of \pkg{data.table} for any data frame-like object. It can turn any vector/matrix/data frame into a time-aware indexed series or frame and perform operations such as lagging, differencing, scaling or centering, encompassing and enhancing core manipulation functionality of \pkg{plm}, \pkg{fixest} and \pkg{xts}. It also performs fast (grouped, weighted) statistical computations along the columns of matrix-like objects, complementing and enhancing \pkg{matrixStats} and \pkg{Rfast}. Its low-level vectorizations and workhorse algorithms are accessible at the \proglang{R} and C-levels, unlike \pkg{data.table}, where most vectorizations and algorithms are internal. It also supports variable labels and intelligently preserves attributes of all objects, complementing \pkg{labelled}. It provides general (recursive) tools to deal with nested data, enhancing \pkg{tidyr}, \pkg{purrr} and \pkg{rrapply}. Finally, it provides a small but consistent and powerful set of descriptive statistical tools, yielding sufficient detail for most data exploration purposes and requiring users to invoke packages like \pkg{DescTools} or \pkg{survey} only for specific statistics. In summary, \pkg{collapse} is a foundation package for statistical computing and data manipulation in \proglang{R} that enhances and integrates seamlessly with the \proglang{R} ecosystem while being outstandingly computationally efficient. A significant benefit is that, rather than piecing together a somewhat fragmented ecosystem oriented at different classes and tasks, core computational tasks can be done with \pkg{collapse}, and easily extended by more specialized packages. This tends to result in \proglang{R} scripts that are shorter, more efficient, and more lightweight in dependencies. \newline
% TODO: Need examples of this

Other programming environments such as \proglang{Python} and \proglang{Julia} now also offer computationally very powerful libraries for tabular data such as \pkg{DataFrames.jl} \citep{jldataframes}, \pkg{Polars} \citep{pypolars} and \pkg{Pandas} \citep{mckinney2010pandas, pypandas}, and supporting numerical libraries such as \pkg{Numpy} \citep{pynumpy} or \pkg{StatsBase.jl} \citep{jlstatsbase}. % \pkg{NaNStatistics.jl} \citep{jlnanstatistics}
In comparison with these, \pkg{collapse} offers a class-agnostic approach bridging the divide between data frames and atomic structures, has more advanced statistical capabilities,\footnote{Such as weighted statistics, including various quantile and mode estimators, support for fully time-aware computations on irregular series/panels, higher order centering, advanced (grouped, weighted, panel-decomposed) descriptive statistics etc., all supporting missing values.} supports recursive operations, variable labels, verbosity for critical operations such as joins, and is extensively globally configurable. In short, it is very utile for complex statistical workflows, rich datasets (e.g., surveys), and for integrating with different parts of the \proglang{R} ecosystem. On the other hand, \pkg{collapse}, for the most part, does not offer a sub-column-level parallel architecture and is thus not highly competitive with top frameworks, including \pkg{data.table}, on aggregating billion-row datasets with few columns\footnote{As can be seen in the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}: \pkg{collapse} is highly competitive on the 10-100 million observations datasets, but deteriorates in performance at larger data sizes (except for joins where it remains competitive). There may be performance improvements for "long data" in the future, but, at present, the treatment of columns as fundamental units of computation is a tradeoff for the highly flexible class-agnostic architecture.}. Its vectorization capabilities are also limited to the statistical functions it provides and not, like \pkg{DataFrames.jl}, to any Julia function. However, as demonstrated below, the possibility of combining vectorized statistical functions also permits calculating more complex statistics in a vectorized way. \newline

@@

\pkg{collapse} is too large and complex to fully present it in a single article or even to present selected topics in depth. The following sections therefore briefly introduce its key components: (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) joins and reshaping; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options (configurability). Section (\ref{sec:bench}) provides a small benchmark, Section (\ref{sec:conclusion}) concludes. For deeper engagement with \pkg{collapse}, the short vignette summarizing available \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{Documentation and Resources} is an excellent starting point.

@@

The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{Fast Statistical Functions}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs} and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R. Specifically, operations such as calculating the mean via the S3 generic \code{fmean()} function are vectorized across columns and groups. They may also involve weights or transformations of the original data. The basic syntax of these functions is

@@

with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e., simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation). The use of single instruction multiple data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.}. The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette (and executed there) demonstrate their basic usage to calculate (column-wise, grouped, weighted) statistics on different objects. As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette} on object handling, these statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix} and \class{data.frame}, which call corresponding C implementations that intelligently preserve object attributes. Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods. Users can also directly call the basis methods in case S3 dispatch does not yield the intended outcome. For example, \code{fmean.default(EuStockMarkets)} computes the mean of the entire matrix. %, and attributes are preserved as much as possible ()

@@

\code{TRA()} is called internally in the \emph{Fast Statistical Functions}, the \code{TRA} argument is passed to \code{FUN}. Thus \code{fmean(x, g, w, TRA = "-")} is equivalent to \code{TRA(x, fmean(x, g, w), "-", g)}. The \code{set} argument can also be passed to \emph{Fast Statistical Functions} to toggle transformation by reference. The following examples demonstrate how this design allows flexible ad-hoc transformations using \proglang{R}'s built-in \code{airquality} dataset with daily measurements for 5 months.

@@

Whereas the \code{g} argument supports ad-hoc grouping with vectors and lists/data frames, the cost of grouping can be optimized by using factors or, even better, \class{GRP} objects, which contain all information for \pkg{collapse}'s vectorized functions to operate across groups efficiently. These objects can be created with \code{GRP()}. The syntax is

@@

\class{GRP} objects make grouped statistical computations in \pkg{collapse} fully programmable. Below, the object is used with the \emph{Fast Statistical Functions} and some utility functions to aggregate the data (with optional weights) efficiently.

@@

This programming access can become very useful. For example, the useR 2022 presentation \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{Slides 18-19} aggregates the \href{https://worldmrio.com/}{EORA Global Supply Chain Database} from the country to the world region level. After defining a single grouping object, a list of value-added shares matrices (VB) and outputs for years 1990-2021, is aggregated with no grouping cost using a single line of code like \code{lapply(VB_list, function(x) x$VB |> fsum(g) |> t() |> fmean(g, x$O) |> t())}. On an M1 Mac using 4 threads, this computation, involving 44.7 million summations and 2.6 million weighted means, takes only 0.33 seconds.\footnote{Another example from my recent work involved numerically optimizing a parameter $a$ in an equation of the form $y_j = \sum_i x_{ij}^a\ \forall j\in J$ where there are $J$ groups (1 million in my case), and the optimal value of $a$ is determined by the proximity of the aggregated vector \textbf{y} to another vector \textbf{z}. Thus each iteration of the numerical routine raises the vector \textbf{x} to a different power ($a$), sums it in 1 million groups ($j$) to generate \textbf{y}, and computes the Euclidean distance to \textbf{z} (using \code{collapse::fdist}). Without grouping objects and vectorization, this would have been difficult to handle within reasonable computing times (a few seconds on the M1).}

@@

the mean of columns \code{mpg}, \code{carb} and \code{hp}, and the weighted mean of \code{qsec}, after subsetting and grouping the data. This code is very fast (especially with many groups) because data does not need to be split by groups at all. There is also no need to call \code{lapply()} inside the \code{across()} statement: \code{fmean.data.frame()} is applied to a subset of the data containing the three columns.\footnote{Internally, the \code{g} argument of the statistical functions is set as a keyword argument by \code{fsummarise/across} and the function is evaluated on a suitable subset of columns. Thus \code{w} becomes the second positional argument...} The \emph{Fast Statistical Functions} also have a method for grouped data, so \code{fsummarise} is not always needed. The following code calculates weighted group means. By default (\code{keep.w = TRUE}) \code{fmean.grouped\_df} also sums the weights in each group.

@@

\subsection{Vectorizations for advanced tasks}

@@

Apart from vectorization, this code avoids 3 intermediate copies: (1) \code{fmean(carb, TRA = "-")} avoids an expanded vector of group means, (2) \code{fsum(mpg, dm\_carb)} uses the weights (\code{w}) argument to \fct{fsum} to avoid materializing a multiplication (as in \code{fsum(mpg * dm\_carb)}), and (3) division by reference (\code{\%/=\%}) avoids allocating an additional vector for the final result. Under the hood, the expression boils down to an (expensive) grouping step, 5 allocations (of which 2 full length), and 6 loops in C to calculate the result. Any modern laptop can calculate 1 million regression slopes in less than 1 second like this. Another neat example, shared by Andrew Ghazi in a recent \href{https://andrewghazi.github.io/posts/collapse_is_sick/sick.html}{blog post}, involved vectorizing an expression to compute the p-value across 300k groups for a simulation study, yielding a 70x performance increase. \newline

\pkg{collapse} also supplies advanced statistics, such as weighted medians and modes. The following example calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}\footnote{\pkg{collapse} calculates weighted quantiles by replacing the sample size with the sum of weights and 1 with the minimum non-zero weight in the respective quantile definition. See \href{https://sebkrantz.github.io/collapse/reference/fquantile.html}{fquantile} for more details.}, and a weighted maximum mode.\footnote{The weighted maximum mode is the largest element with the maximum sum of weights.}

@@

\pkg{collapse} provides a flexible high-performance architecture to perform (\emph{time aware}) computations on time series and panel series. In particular, the user enjoys great flexibility in deciding the desired degree of indexation and mode of computation. It is possible to apply time series and panel data transformations without any indexation by passing individual and/or time identifiers to the respective functions in an ad-hoc fashion or by using \class{indexed\_frame} and \class{indexes\_series} classes, which implement full and deep indexation. Table~\ref{tab:TSfun} summarizes \pkg{collapse}'s time series and panel data architecture.

@@

\emph{Classes, Constructors and Utilities} \\

@@

\emph{Core Time-Based Functions} \\

@@

\emph{Data Transformation Functions with Supporting Methods} \\

@@

\emph{Data Manipulation Functions with Supporting Methods} \\

@@

Time series functions such as \fct{fgrowth} (to compute growth rates) are S3 generic and can be applied to most time series classes. In addition to a \code{g} argument for grouped computation, these functions also have a \code{t} argument for indexation. If \code{t} is a plain numeric vector or a factor, it is coerced to integer and interpreted as time steps\footnote{This is premised on the observation that the most common form of temporal identifier is a numeric variable denoting calendar years.}. If \code{t} is a numeric time object (e.g., \class{Date}, \class{POSIXct}, etc.), then it is internally passed through \code{timeid()} which computes the greatest common divisor (GCD) and generates an integer time-id. For the GCD approach to work, \code{t} must have an appropriate class, e.g., for monthly/quarterly data, \code{zoo::yearmon()/zoo::yearqtr()} should be used instead of \class{Date} or \class{POSIXct}. % The following provides a basic example:

@@

\pkg{collapse}'s list processing toolkit provides further useful functions such as \fct{t\_list} to turn lists of lists inside out, \fct{has\_elem} to check for the existence of elements, \fct{ldepth} to return the maximum level of recursion, and \fct{is\_unlistable} to check whether a list has atomic elements in all final nodes. A non-recursive and class-agnostic \fct{rowbind} function also exists to efficiently bind lists of data frame-like objects (like \code{data.table::rbindlist}).

>>>>

%% - \Abstract{} almost as usual
\Abstract{
\pkg{collapse} is a large C/C++-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and programming efficiency. It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g., \class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling seamless integration with large parts of the \proglang{R} ecosystem. This article introduces the package's key components and design principles in a structured way, supported by a rich set of examples. A small benchmark demonstrates its computational performance.
%  This short article illustrates how to write a manuscript for the
%  \emph{Journal of Statistical Software} (JSS) using its {\LaTeX} style files.
%  Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
%  it is recommended to keep the {\LaTeX} code as simple as possible,
%  i.e., avoid inclusion of packages/commands that are not necessary.
%  For outlining the typical structure of a JSS article some brief text snippets
%  are employed that have been inspired by \cite{Zeileis+Kleiber+Jackman:2008},
%  discussing count data regression in \proglang{R}. Editorial comments and
%  instructions are marked by vertical bars.
}

@@

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}}\footnote{Website: https://sebkrantz.github.io/collapse/} is a large C/C++-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions. Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently. It also offers vectorizations for advanced operations such as weighted statistics (including mode and quantiles), functions and classes for fully indexed (time-aware) computations on time series and panel data, recursive (list-processing) tools to deal with nested data and advanced descriptive statistical tools. This functionality is supported by efficient algorithms for intensive operations like grouping, unique values, matching, ordering, etc., tailored to \proglang{R}'s data structures, and powerful data manipulation functions. The package also supplies many features for memory efficient \proglang{R} programming, such as data transformation and math by reference, and aversion of logical vectors. \pkg{collapse} is class-agnostic, i.e., it provides most statistical operations for atomic vectors, matrices, and data frames/lists, and seamlessly supports key variants of these objects used in the \proglang{R} ecosystem (e.g., \class{tibble}, \class{data.table}, \class{sf}, \class{xts}, \class{pdata.frame}). It is globally and interactively configurable, which includes setting different defaults for key function arguments (such as \code{na.rm} arguments to statistical functions, default \code{TRUE}), and modifying the package namespace itself.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse}, but can be interactively modified to overwrite key functions like \code{unique}, \code{match}, \code{\%in\%}, \code{table}, \code{subset}, \code{mutate}, \code{summarise} etc. with much faster \pkg{collapse} equivalents. See Section~\ref{sec:glob_opt}.} \newline

What is the purpose of combining all of this in a package? The short answer is to make computations in \proglang{R} as flexible and powerful as possible. The more elaborate answer is to (1) facilitate complex data transformation, exploration, and computing tasks in \proglang{R}; (2) increase performance and parsimony by avoiding \proglang{R}-level repetition;\footnote{Such as applying \proglang{R} functions across columns or split-apply-combine computing to apply functions across groups or other divisions of data.} (3) increase the memory efficiency and flexibility of \proglang{R} programs;\footnote{E.g., by avoiding object conversions and the need for certain classes to do certain things, such as converting to data frame or \class{data.table} to do something "by groups" and then convert back to matrix to continue with linear algebra, and in general to reduce the need for metaprogramming.} and (4) to create a new foundation package for statistics and data manipulation in \proglang{R} that implements successful ideas developed in the \proglang{R} ecosystem and other programming environments such as \proglang{Python} or \proglang{STATA} \citep{STATA}, including some new ideas, in a stable, high performance, and broadly compatible manor.\footnote{Examples of such ideas are \pkg{tidyverse} syntax, vectorized aggregations (\pkg{data.table}), data transformation by reference (\proglang{Python}, \pkg{pandas}), vectorized and verbose joins (\pkg{polars}, \proglang{STATA}), indexed time series and panel data (\pkg{xts}, \pkg{plm}), summary statistics for panel data (\proglang{STATA}), reshaping labelled data (myself) etc...} \newline

@@

\pkg{collapse} relates to and integrates key elements from these projects. It offers \pkg{tidyverse}-like data manipulation at the speed and stability of \pkg{data.table} for any data frame-like object. It can turn any vector/matrix/data frame into a time-aware indexed series or frame and perform operations such as lagging, differencing, scaling or centering, encompassing and enhancing core manipulation functionality of \pkg{plm}, \pkg{fixest}, and \pkg{xts}. It also performs fast (grouped, weighted) statistical computations along the columns of matrix-like objects, complementing and enhancing \pkg{matrixStats} and \pkg{Rfast}. Its low-level vectorizations and workhorse algorithms are accessible at the \proglang{R} and C-levels, unlike \pkg{data.table}, where most vectorizations and algorithms are internal. It also supports variable labels and intelligently preserves attributes of all objects, complementing \pkg{labelled}. It provides general (recursive) tools to deal with nested data, enhancing \pkg{tidyr}, \pkg{purrr}, and \pkg{rrapply}. Finally, it provides a small but consistent and powerful set of descriptive statistical tools, yielding sufficient detail for most data exploration purposes, requiring users to invoke packages like \pkg{DescTools} or \pkg{survey} only for specific statistics. In summary, \pkg{collapse} is a foundation package for statistical computing and data manipulation in \proglang{R} that enhances and integrates seamlessly with the \proglang{R} ecosystem while being outstandingly computationally efficient. A significant benefit is that, rather than piecing together a somewhat fragmented ecosystem oriented at different classes and tasks, core computational tasks can be done with \pkg{collapse}, and easily extended by more specialized packages. This tends to result in \proglang{R} scripts that are shorter, more efficient, and more lightweight in dependencies. \newline
% TODO: Need examples of this

Other programming environments such as \proglang{Python} and \proglang{Julia} now also offer computationally very powerful libraries for tabular data such as \pkg{DataFrames.jl} \citep{jldataframes}, \pkg{Polars} \citep{pypolars}, and \pkg{Pandas} \citep{mckinney2010pandas, pypandas}, and supporting numerical libraries such as \pkg{Numpy} \citep{pynumpy}, or \pkg{StatsBase.jl} \citep{jlstatsbase}. % \pkg{NaNStatistics.jl} \citep{jlnanstatistics}
In comparison with these, \pkg{collapse} offers a class-agnostic approach bridging the divide between data frames and atomic structures, has more advanced statistical capabilities,\footnote{Such as weighted statistics, including various quantile and mode estimators, support for fully time-aware computations on irregular series/panels, higher order centering, advanced (grouped, weighted, panel-decomposed) descriptive statistics etc., all supporting missing values.} supports recursive operations, variable labels, verbosity for critical operations such as joins, and is extensively globally configurable. In short, it is very utile for complex statistical workflows, rich datasets (e.g., surveys), and for integrating with different parts of the \proglang{R} ecosystem. On the other hand, \pkg{collapse}, for the most part, does not offer a sub-column-level parallel architecture and is thus not highly competitive with top frameworks, including \pkg{data.table}, on aggregating billion-row datasets with few columns.\footnote{As can be seen in the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}: \pkg{collapse} is highly competitive on the 10-100 million observations datasets, but deteriorates in performance at larger data sizes (except for joins where it remains competitive). There may be performance improvements for "long data" in the future, but, at present, the treatment of columns as fundamental units of computation is a tradeoff for the highly flexible class-agnostic architecture.} Its vectorization capabilities are also limited to the statistical functions it provides and not, like \pkg{DataFrames.jl}, to any Julia function. However, as demonstrated in Section~\ref{ssec:vfat}, vectorized statistical functions can be combined to calculate more complex statistics in a vectorized way. \newline

@@

\pkg{collapse} is too large and complex to fully present it in a single article, or even to present selected topics in depth. The following sections therefore briefly introduce its key components: (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) joins and reshaping; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options. Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes. For deeper engagement with \pkg{collapse}, the short vignette summarizing available \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation and resources} is an excellent starting point.

@@

The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs} and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R. Specifically, operations such as calculating the mean via the S3 generic \code{fmean()} function are vectorized across columns and groups. They may also involve weights or transformations of the original data. The basic syntax of these functions is

@@

with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e., simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading.\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation). The use of single instruction multiple data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.} The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette (and executed there) demonstrate their basic usage to calculate (column-wise, grouped, weighted) statistics on different objects. As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding C implementations that intelligently preserve object attributes. Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods. Users can also directly call the basis methods in case S3 dispatch does not yield the intended outcome. For example, \code{fmean.default(EuStockMarkets)} computes the mean of the entire matrix. %, and attributes are preserved as much as possible ()

@@

\code{TRA()} is called internally in the \emph{Fast Statistical Functions}, the \code{TRA} argument is passed to \code{FUN}. Thus \code{fmean(x, g, w, TRA = "-")} is equivalent to \code{TRA(x, fmean(x, g, w), "-", g)}. The \code{set} argument can also be passed to \emph{Fast Statistical Functions} to toggle transformation by reference. The following examples demonstrate how this design allows flexible ad-hoc transformations using \proglang{R}'s built-in \code{airquality} dataset with daily measurements in New York from May to September 1973.

@@

Whereas the \code{g} argument supports ad-hoc grouping with vectors and lists/data frames, the cost of grouping can be optimized by using factors or, even better, \class{GRP} objects, which readily contain all information \pkg{collapse}'s vectorized statistical functions might require to operate across groups. These objects can be created with \code{GRP()}. The syntax is

@@

\class{GRP} objects make grouped statistical computations in \pkg{collapse} fully programmable. Below, the object is used with the \emph{Fast Statistical Functions} and some utility functions to efficiently aggregate data (with optional frequency weights).

@@

This programming access can become very useful. For example, the useR 2022 presentation \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{Slides 18-19} aggregates the \href{https://worldmrio.com/}{EORA Global Supply Chain Database} from the country to the world region level. After defining a single grouping object, a list of value-added shares matrices (VB) and outputs for years 1990-2021, is aggregated with no grouping cost using a single line of code like \code{lapply(VB_list, function(x) x$VB |> fsum(g) |> t() |> fmean(g, x$O) |> t())}. On an M1 Mac using 4 threads, this computation, involving 44.7 million summations and 2.6 million weighted means, takes only 0.33 seconds.\footnote{Another example from my recent work involved numerically optimizing a parameter $a$ in an equation of the form $y_j = \sum_i x_{ij}^a\ \forall j\in J$ where there are $J$ groups (1 million in my case), and the optimal value of $a$ is determined by the proximity of the aggregated vector \textbf{y} to another vector \textbf{z}. Thus each iteration of the numerical routine raises the vector \textbf{x} to a different power ($a$), sums it in 1 million groups ($j$) to generate \textbf{y}, and computes the Euclidean distance to \textbf{z} (using \code{collapse::fdist}). Without grouping objects and vectorization, this would have been difficult to handle within reasonable computing times (of a few seconds on the M1).}

@@

the mean of columns \code{mpg}, \code{carb} and \code{hp}, and the weighted mean of \code{qsec}, after subsetting and grouping the data. This code is very fast (especially with many groups) because data does not need to be split by groups at all. There is also no need to call \code{lapply()} inside the \code{across()} statement: \code{fmean.data.frame()} is applied to a subset of the data containing the three columns.\footnote{Internally, the \code{g} argument of the statistical functions is set as a keyword argument by \code{fsummarise/across} and the function is evaluated on a suitable subset of columns. Thus \code{w} becomes the second positional argument...} The \emph{Fast Statistical Functions} also have a method for grouped data, so \code{fsummarise} is not always needed. The following code calculates weighted group means. By default (\code{keep.w = TRUE}) \code{fmean.grouped\_df} also sums the weights in each group.\footnote{\class{grouped\_df} methods in \pkg{collapse} support grouped data created with either \fct{fgroup\_by} or \fct{dplyr::fgroup\_by}. The latter requires an additional C routine to convert the grouping object to a \class{GRP} object, and is thus much less efficient.}

@@

\subsection{Vectorizations for advanced tasks} \label{ssec:vfat}

@@

Apart from vectorization, this code avoids 3 intermediate copies: (1) \code{fmean(carb, TRA = "-")} avoids an expanded vector of group means, (2) \code{fsum(mpg, dm\_carb)} uses the weights (\code{w}) argument to \fct{fsum} to avoid materializing a multiplication (as in \code{fsum(mpg * dm\_carb)}), and (3) division by reference (\code{\%/=\%}) avoids allocating an additional vector for the final result. Under the hood, the expression boils down to an (expensive) grouping step, 5 allocations (of which 2 full length), and 6 loops in C to calculate the result. Any modern laptop can calculate 1 million regression slopes in less than 1 second like this. Another neat example, shared by Andrew Ghazi in a recent \href{https://andrewghazi.github.io/posts/collapse_is_sick/sick.html}{blog post}, vectorizes an expression to compute the p-value across 300k groups for a simulation study, yielding a 70x performance increase. \newline

\pkg{collapse} also vectorizes advanced statistics, such as weighted medians and modes. The following example calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}.\footnote{\pkg{collapse} calculates weighted quantiles by replacing the sample size with the sum of weights and 1 with the minimum non-zero weight in the respective quantile definition. See \href{https://sebkrantz.github.io/collapse/reference/fquantile.html}{fquantile} for more details.} and a weighted maximum mode.\footnote{The weighted maximum mode is the largest element with the maximum sum of weights.}

@@

\pkg{collapse} also provides a flexible high-performance architecture to perform (\emph{time aware}) computations on time series and panel series. In particular, the user enjoys great flexibility in deciding the desired degree of indexation and mode of computation. It is possible to apply time series and panel data transformations without any indexation by passing individual and/or time identifiers to the respective functions in an ad-hoc fashion, or by using \class{indexed\_frame} and \class{indexes\_series} classes, which implement full and deep indexation. Table~\ref{tab:TSfun} summarizes \pkg{collapse}'s time series and panel data architecture.

@@

\emph{Classes, constructors and utilities} \\

@@

\emph{Core time-based functions} \\

@@

\emph{Data transformation functions with supporting methods} \\

@@

\emph{Data manipulation functions with supporting methods} \\

@@

Time series functions such as \fct{fgrowth} (to compute growth rates) are S3 generic and can be applied to most time series classes. In addition to a \code{g} argument for grouped computation, these functions also have a \code{t} argument for indexation. If \code{t} is a plain numeric vector or a factor, it is coerced to integer and interpreted as time steps.\footnote{This is premised on the observation that the most common form of temporal identifier is a numeric variable denoting calendar years.} If \code{t} is a numeric time object (e.g., \class{Date}, \class{POSIXct}, etc.), then it is internally passed through \code{timeid()} which computes the greatest common divisor (GCD) and generates an integer time-id. For the GCD approach to work, \code{t} must have an appropriate class, e.g., for monthly/quarterly data, \code{zoo::yearmon()/zoo::yearqtr()} should be used instead of \class{Date} or \class{POSIXct}. % The following provides a basic example:

@@

\pkg{collapse}'s list processing toolkit provides further useful functions such as \fct{t\_list} to turn lists of lists inside out, \fct{has\_elem} to check for the existence of elements, \fct{ldepth} to return the maximum level of recursion, and \fct{is\_unlistable} to check whether a list has atomic elements in all final nodes. A non-recursive and class-agnostic \fct{rowbind} function also exists to efficiently bind lists of data frame-like objects (like \code{data.table::rbindlist()}).
